---
title: "class05_data_visualization"
author: "James Woolley, A16440072"
format: pdf
editor: visual
---

## using GGPLOT

The ggplot2 package needs to be installed here as it does not come with R "out of the box". Used the `install.packages()` function to do it.

To use ggplot, you need to load it before using any of the functions it comes with. Do this with the `library()` function.
```{r}
head(cars)
```
```{r}
library(ggplot2)
ggplot(cars) +
  aes(x=speed,y=dist) +
  geom_point() +
  labs(title="Distance to Stop at Differenrt Speeds of Cars",
       x="Speed (MPH)",
       y="Distance (Ft)",
       subtitle="This is a plot with a line of best fit",
       caption="Dataset: Cars")+
  geom_smooth(method = "lm",se=FALSE)
```
All ggplot figures have at least 3 things. 1- data 2- aesthetic mapping 3- geoms

There are lots of graphing systems in R, including one that comes with it.
```{r}
plot(cars)
```

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
```{r}
nrow(genes)
colnames(genes)
ncol(genes)
table(genes$State)
round( table(genes$State)/nrow(genes) * 100, 2 )
```
```{r}
library(ggplot2)
p <- ggplot(genes)+
  aes(x=Condition1,y=Condition2, col=State) +
  geom_point()+
  scale_colour_manual( values=c("orange","green","white") )+
   labs(title="Gene Expression Changes Upon Drug Treatment",
       x="Control (no drug)",
       y="Drug Treatment")
p
```



```{r}
allstates <- genes$State
nstates <- table(allstates)
nstates/nrow(genes)
```

